Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-17 forecastflows.tex
making some progress

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. % jdavies15april: could refer to the 'pre-registration' movement here ...
But this idealized method often does not apply to the reality of ecological research, and most of macroecological insights emerge from exploring patterns in data. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.

This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means
that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models?
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2?
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig. \ref{fig:workflowy}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question. The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig. \ref{fig:workflowy}). This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo)
Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, % structural degeneracies in the model might become apparent, 
this step might reveal that some parameters are highly non-identifiable,
flagging % not sure about this word
the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?

@@

% vvdm07april: need to work on a better intro paragraph
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....

>>>>

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses---following the principles of pre-registration, where hypotheses and methodology are defined prior to data collection.  % jdavies15april: could refer to the 'pre-registration' movement here ...
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means... [FLAG]
that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models? 
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig. \ref{fig:workflowy}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question. The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig. \ref{fig:workflowy}). % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo), vvdm16april: reformulate:
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting. 
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to come out with a research question.

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable,
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*? [FLAG]

@@

Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
If a model becomes too complex, it may become a black box, and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... [FLAG: agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section...]
